Stagnating State Funding for Higher Education and its Effect on Faculty Composition at US Universities

### 2023-06-02 text/sections/00-abstract.tex
Adjusted Table 3 + fixed spelling mistakes.

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while at the same time states have limited how much they fund higher education.
This analysis relates the changes the faculty composition at US public universities to the fall in state funding for higher education, using a shift-share approach to identify causal effects.
A fall of $-10$\% in university funding, via a state funding shock, decreases the number of assistant professors per student by $-1.4$\% and full professors by $-1.2$\%, yet increases the number of lecturers per student by 4.4\%; long-run estimates show that the faculty composition is affected both in the same year as the funding shock, and two to three years following.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, with suggestive evidence showing that the faculty changes arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

>>>>

US universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while at the same time states have limited how much they fund higher education.
This analysis relates the changes the faculty composition at US public universities to the fall in state funding for higher education, using a shift-share approach to instrument for state funding and identify causal effects.
A fall of $-10$\% in university funding, via a state funding shock, decreases the number of assistant professors per student by $-1.4$\% and full professors by $-1.2$\%, yet increases the number of lecturers per student by 4.4\%; long-run estimates show that the faculty composition is affected both in the same year as the funding shock, and two to three years following.
Analysis of all the professors at Illinois public universities shows that incumbent professors are not affected by substantial funding shocks over 2011-2021, with suggestive evidence showing that the faculty changes arose by limiting the hiring of new professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and raise questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

### 2023-06-02 text/sections/02-data.tex
Fixed spelling.
Adjusted Table 3 + fixed spelling mistakes.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw NCES data on 2018-2021 for all relevant variables.
    Figures for enrolment, faculty counts and salaries come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.

@@

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{IPEDS Summary Statistics, Public Universities Panel 1987--2021}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources, so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.

>>>>

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw NCES data on 2018-2021 for all relevant variables.
    Figures for enrolment and faculty counts come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.

@@

\begin{table}[h!]
    \singlespacing
    %\centering
    \caption{IPEDS Summary Statistics, Public Universities Panel 1987--2021}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, but only provides aggregated information within a university.\footnote{
    IPEDS additionally gives a mean salary measure for faculty by rank (inconsistently) for each university.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period compared to summary statistics provided in \citealt{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources (including those calculated for in Illinois in the IBHED data later in this section) so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
These publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.
